---
title: "Weekly residence model"
subtitle: "Quasibinomial"
author: "Mike O'Brien"
date: "9/1/2021"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table); library(mgcv); library(itsadug)

basswl <- fread('bassweek.csv')
basswl[, Origin := factor(Origin)]
basswl[, transmitter := factor(transmitter)]

# Need to label where each fish's time series starts for the AR1 model
basswl[, ar_start := fifelse(time == 1, T, F)]
basswl[, trials := 7]
```

The following fits models with autocorrelation parameters ($\rho$) from
0 to 0.999 and retains the AIC returned.

Note the use of `scale = -1`. This this tricks `mgcv::bam` into fitting
a scale parameter, meaning that we are technically fitting a
quasibinomial model. If you use the `quasibinomial` function directly,
AIC is not reported since it technically is a quasilikelihood that
shouldn't be compared to other quasilikelihoods. The R creators are
purists and don't want you to see that the likelihood is still
calculated behind the curtain and that AIC can still be pulled out.

I also increased k in `s(WE)` to 28 since the cyclic spline seems to
take 2 degrees of freedom. Using 28 rather than 27 gives us the max of
26 knots that we were looking for.

```{r eval=FALSE}
search <- sapply(seq(0, 0.999, 0.001), function(x){
  AIC(mgcv::bam(cbind(freq, trials) ~
                  Origin + 
                  s(SST, bs='tp', k = 11) +
                  s(WE, bs='cc', k = 28) +
                  s(CHLA, bs= 'tp', k = 6) +
                  s(transmitter, bs = 're'),
                rho = x, AR.start = ar_start,
                family = binomial, discrete = T,
                
                scale = -1,
                
                data=basswl))
})
```

It comes out that $\rho$ = 0.865 is the optimal value according to AIC,
so if you want to run it yourself, I'd change the sequence range to
something like `seq(0.85, 0.87, 0.001)` to make it go faster.

```{r}
search <- sapply(seq(0.86, 0.87, 0.001), function(x){
  AIC(mgcv::bam(cbind(freq, trials) ~
                  Origin + 
                  s(SST, bs='tp', k = 11) +
                  s(WE, bs='cc', k = 28) +
                  s(CHLA, bs= 'tp', k = 6) +
                  s(transmitter, bs = 're'),
                rho = x, AR.start = ar_start,
                family = binomial, discrete = T,
                
                scale = -1,
                
                data=basswl))
})
```

Return the best value of $\rho$ (note that I did change the sequence
range once I had an idea of where $\rho$ was to make it run faster):

```{r}
seq(0.86, 0.87, 0.001)[which.min(search)]
```

And now, run the model:

```{r}
model <- bam(cbind(freq, trials) ~
               Origin +
               s(SST, bs='tp', k = 11) +
               s(WE, bs='cc', k = 28) +
               s(CHLA, bs= 'tp', k = 6) +
               s(transmitter, bs = 're'),
             rho = 0.865, AR.start = basswl$ar_start,
             family = binomial, discrete = T, 
             scale = -1, data=basswl)

summary(model)
```

Note that "Scale est." is much less than 1. This indicates that that the
data are, indeed, underdispersed. If you run this using the binomial
family without `scale = -1`, you will see that "Scale est." is fixed at
1.

The adjusted R^2^ and deviance explained are obviously pretty funky, but
I think I can calculate those by hand later.

CHLA is basically linear, so re-running without a smooth for that term:

```{r}
model <- bam(cbind(freq, trials) ~
               Origin +
               s(SST, bs='tp', k = 11) +
               s(WE, bs='cc', k = 28) +
               CHLA +
               s(transmitter, bs = 're'),
             rho = 0.865, AR.start = basswl$ar_start,
             family = binomial, discrete = T, 
             scale = -1, data=basswl)

summary(model)
```

```{r}
plot(model, scale = 0, pages = 1)
```

You can see from the random subset of fish below that there's still some
autocorrelation, but we've done the best we probably can!

```{r}
acf_resid(model, split_pred = 'AR.start', n = 6)
```

Looks like the Delaware and Hudson may be different...

```{r}
plot_parametric(model,
                pred = list(Origin = levels(basswl$Origin)),
                parametricOnly = T)
```

...and they seem to be!

```{r}
wald_gam(model,
         comp = list(Origin = levels(basswl$Origin)),
         print.output = T)
```
