---
title: "D50 GAM with all variables"
author: "Mike O'Brien"
date: "Jan 4, 2019"
output: html_notebook
---
```{r setup, echo = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

The first two GAMM iterations used 1) predictors that were recorded in situ by each receiver and 2) predictios that can be accessed through freely-available ocean observing system (OOS) outlets. Now, we're going to move onto throwing everything into the kitchen sink. The only new variable here will be `dt`, which is simply SST - BWT.

## Variable selection
Pull out the predictor variables, keeping only the dates that have observations for them all.

```{r}
data <- readRDS('data and imports/rangetest_no_outliers.RDS')

wea <- data[data$array == 'MD WEA',
            names(data) %in% c('average_temperature', 'average_noise', 'tilt_angle',
                               'dt', 'sst', 'wdir', 'wspd', 'gst', 'wvht', 'dpd',
                               'apd', 'mwd', 'pres', 'atmp', 'wtmp')]
wea <- wea[complete.cases(wea),]
summary(wea)
```

Now for the pairs plot.

```{r}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...){
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8 / strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}

pairs(wea, lower.panel = panel.smooth, upper.panel = panel.cor, gap = 0,
      row1attop = F)
```

After tacking on $\Delta T$ ("dt") and the in-situ variables, we can see that BWT ("average_temperature") is correlated with, SST (0.85), $\Delta T$ (0.70), and the buoy-recorded temperature measures ("wtemp": 0.84; "atmp": 0.77). In addition, noise ("average_noise") is correlated with wave height ("Wvht") and average wave period ("apd").

Now to run our favorite PCA.

```{r}
pca <- vegan::rda(wea, scale = T)
biplot(pca, type = c('text', 'points'))
```

As we saw before, the temperature metrics are all collinear. Since I have a hunch that **$\Delta T$** will explain much more of what's going on, I'm dropping the other temperature measures. 

I'll also drop wind gust speed in favor of **wind speed**, as the aggregate measure of wind is more likely to affect detection distance than an acute metric.

```{r}
wea_dropvar <- wea[, !names(wea) %in% c('average_temperature', 'atmp', 'sst',
                                        'wtmp', 'gst')]
pca2 <- vegan::rda(wea_dropvar, scale = T)
biplot(pca2, type = c('text', 'points'))
```

At this point, I feel like there's a pretty good spread. I will probably have to come back and think about selecting one of *mwd*/*wdir* and *dpd*/*apd*/*tilt_angle*.

## Modeling
$\Delta T$, tilt angle, and noise were recorded individually at each array, so I'll allow them to vary randomly with array.

```{r}
mod_data <- data[, names(data) %in% c('date', 'array', 'd50_adj', 'average_noise',
                                      'tilt_angle', 'dt', 'wdir', 'wspd', 'wvht',
                                      'dpd', 'apd', 'mwd', 'pres')]
mod_data <- mod_data[complete.cases(mod_data[, !names(mod_data) %in% 'date']),]

library(mgcv)
mod <- gam(d50_adj ~ s(dt, array, bs = 're') + s(dt, bs = 'ts') +
             s(average_noise, array, bs = 're') + s(average_noise, bs = 'ts') +
             s(tilt_angle, array, bs = 're') + s(tilt_angle, bs = 'ts') +
             s(wdir, bs = 'ts')+
             s(wspd, bs = 'ts') +
             s(wvht, bs = 'ts') +
             s(dpd, bs = 'ts') +
             s(apd, bs = 'ts') +
             s(mwd, bs = 'ts') +
             s(pres, bs = 'ts'),
           data = mod_data,
           family = Gamma(),
           method = 'REML',
           verbosePQL = F)

summary(mod)
```

Removing the random slope of $\Delta T$ by array.

```{r}
mod2 <- gam(d50_adj ~ s(dt, bs = 'ts') +
             s(average_noise, array, bs = 're') + s(average_noise, bs = 'ts') +
             s(tilt_angle, array, bs = 're') + s(tilt_angle, bs = 'ts') +
             s(wdir, bs = 'ts')+
             s(wspd, bs = 'ts') +
             s(wvht, bs = 'ts') +
             s(dpd, bs = 'ts') +
             s(apd, bs = 'ts') +
             s(mwd, bs = 'ts') +
             s(pres, bs = 'ts'),
           data = mod_data,
           family = Gamma(),
           method = 'REML',
           verbosePQL = F)

summary(mod2)
```

And now random slope of tilt angle by array.

```{r}
mod3 <- gam(d50_adj ~ s(dt, bs = 'ts') +
             s(average_noise, array, bs = 're') + s(average_noise, bs = 'ts') +
             s(tilt_angle, bs = 'ts') +
             s(wdir, bs = 'ts')+
             s(wspd, bs = 'ts') +
             s(wvht, bs = 'ts') +
             s(dpd, bs = 'ts') +
             s(apd, bs = 'ts') +
             s(mwd, bs = 'ts') +
             s(pres, bs = 'ts'),
           data = mod_data,
           family = Gamma(),
           method = 'REML',
           verbosePQL = F)

summary(mod3)
```

Random effects seem sorted. Drop wind and wave directions.

```{r}
mod4 <- gam(d50_adj ~ s(dt, bs = 'ts') +
             s(average_noise, array, bs = 're') + s(average_noise, bs = 'ts') +
             s(tilt_angle, bs = 'ts') +
             s(wspd, bs = 'ts') +
             s(wvht, bs = 'ts') +
             s(dpd, bs = 'ts') +
             s(apd, bs = 'ts') +
             s(pres, bs = 'ts'),
           data = mod_data,
           family = Gamma(),
           method = 'REML',
           verbosePQL = F)

summary(mod4)
```

Remove average wave period.

```{r}
mod5 <- gam(d50_adj ~ s(dt, bs = 'ts') +
             s(average_noise, array, bs = 're') + s(average_noise, bs = 'ts') +
             s(tilt_angle, bs = 'ts') +
             s(wspd, bs = 'ts') +
             s(wvht, bs = 'ts') +
             s(dpd, bs = 'ts') +
             s(pres, bs = 'ts'),
           data = mod_data,
           family = Gamma(),
           method = 'REML',
           verbosePQL = F)

summary(mod5)
```

Drop pressure.

```{r}
mod6 <- gam(d50_adj ~ s(dt, bs = 'ts') +
             s(average_noise, array, bs = 're') + s(average_noise, bs = 'ts') +
             s(tilt_angle, bs = 'ts') +
             s(wspd, bs = 'ts') +
             s(wvht, bs = 'ts') +
             s(dpd, bs = 'ts'),
           data = mod_data,
           family = Gamma(),
           method = 'REML',
           verbosePQL = F)

summary(mod6)
```

Everything is significant now, but the dominant wave period seems like it may have a linear response. Check the plots.

```{r}
plot(mod6, pages = 2, scale = 0)
```

Both dominant wave period and wind speed look linear! Change around the model.

```{r}
mod7 <- gam(d50_adj ~ s(dt, bs = 'ts') +
             s(average_noise, array, bs = 're') + s(average_noise, bs = 'ts') +
             s(tilt_angle, bs = 'ts') +
             wspd +
             s(wvht, bs = 'ts') +
             dpd,
           data = mod_data,
           family = Gamma(),
           method = 'REML',
           verbosePQL = F)

summary(mod7)
```

```{r}
plot(mod7, pages = 1, scale = 0)
```


```{r}
marginal_plot <- function(variable){
  newdata_ifelse <- function(x){
    if(variable == x){
      seq(min(mod_data[[variable]]), max(mod_data[[variable]]), length = 100)
    } else{
      mean(mod_data[[x]])
    }
  }
  
  newdata <- data.frame(dt = newdata_ifelse('dt'),
                        average_noise = newdata_ifelse('average_noise'),
                        tilt_angle = newdata_ifelse('tilt_angle'),
                        wvht = newdata_ifelse('wvht'),
                        wspd = newdata_ifelse('wspd'),
                        dpd = newdata_ifelse('dpd'),
                        array = 'MD WEA')
  
  pred <- predict(mod7, newdata, type = 'response',
                  exclude = grep('array', row.names(summary(mod7)$s.table),
                                 value = T))
  pred_se <- predict(mod7, newdata, type = 'link', se = T,
                     exclude = grep('array', row.names(summary(mod7)$s.table),
                                    value = T))
  
  values <- data.frame(resp = pred,
                       new_vals = newdata[[variable]],
                       UCI = 1/(pred_se$fit - 2 * pred_se$se.fit),
                       LCI = 1/(pred_se$fit + 2 * pred_se$se.fit))

  plot(resp ~ new_vals, data = values, type = 'n',
       ylim = c(min(LCI), max(UCI)),
       xlab = variable, ylab = 'D50')
  lines(resp ~ new_vals, data = values)
  lines(LCI ~ new_vals, data = values, lty = 2)
  lines(UCI ~ new_vals, data = values, lty = 2)
  rug(mod_data[[variable]])
}

par(mfrow = c(2, 3))
for(i in c('wspd', 'dpd', 'dt', 'average_noise', 'tilt_angle', 'wvht')) marginal_plot(i)
```


```{r}
p_mod <- predict(mod7, type = 'response',
                 # next line removes random effects (anything with 'array')
                 exclude = grep('array', row.names(summary(mod7)$s.table),
                                value = T))

p_mod_se <- predict(mod7, type = 'link', se = T,
                    exclude = grep('array', row.names(summary(mod7)$s.table),
                                   value = T))

library(dplyr)
p_mod <- cbind(data.frame(mod_data), data.frame(response_fit = p_mod), p_mod_se)
p_mod <- p_mod %>% 
  # This calculates the lower and upper CIs and transforms them to the response scale
  mutate(LCI = 1/(fit + 2 * se.fit),
         UCI = 1/(fit - 2 * se.fit))

p <- p_mod %>%
  select(date, array, Observed = d50_adj, Estimated = response_fit, LCI, UCI) %>%
  # use the time period where all predictors are available
  filter(date <= '2018-08-19') %>%
  tidyr::gather(key = 'key', value = 'val', -date, -array, -LCI, -UCI)

library(ggplot2)
ggplot() +
  geom_ribbon(data = p[p$key == 'Estimated',],
              aes(x = date, ymin = LCI, ymax =  UCI),
              fill = 'gray', alpha = 0.5) +
  geom_line(data = p[p$key %in% c('Observed', 'Estimated'),],
            aes(x = date, y = val, linetype = key)) +
  labs(x=NULL, y = 'Estimated D50', linetype = NULL) +
  facet_wrap(~array, nrow = 2) +
  theme_bw()
```






## Interaction between sst and bwt?

```{r}
mod_data2 <- data[, names(data) %in% c('date', 'array', 'd50_adj', 'average_noise',
                                      'tilt_angle', 'dt', 'wdir', 'wspd', 'wvht',
                                      'dpd', 'apd', 'mwd', 'pres', 'sst', 'average_temperature')]
mod_data2 <- mod_data2[complete.cases(mod_data2[, !names(mod_data2) %in% 'date']),]


mod8 <- gam(d50_adj ~ wpd+
              s(sst, average_temperature, bs = 'ts') +
             s(average_noise, array, bs = 're') + s(average_noise, bs = 'ts') +
             s(wvht, bs = 'ts'),
           data = mod_data2,
           family = Gamma(),
           method = 'REML',
           verbosePQL = F)

summary(mod8)
```

```{r}
p_mod <- predict(mod8, type = 'response',
                 # next line removes random effects (anything with 'array')
                 exclude = grep('array', row.names(summary(mod8)$s.table),
                                value = T))

p_mod_se <- predict(mod8, type = 'link', se = T,
                    exclude = grep('array', row.names(summary(mod8)$s.table),
                                   value = T))

library(dplyr)
p_mod <- cbind(data.frame(mod_data), data.frame(response_fit = p_mod), p_mod_se)
p_mod <- p_mod %>% 
  # This calculates the lower and upper CIs and transforms them to the response scale
  mutate(LCI = 1/(fit + 2 * se.fit),
         UCI = 1/(fit - 2 * se.fit))

p <- p_mod %>%
  select(date, array, Observed = d50_adj, Estimated = response_fit, LCI, UCI) %>%
  # use the time period where all predictors are available
  filter(date <= '2018-08-19') %>%
  tidyr::gather(key = 'key', value = 'val', -date, -array, -LCI, -UCI)

library(ggplot2)
ggplot() +
  geom_ribbon(data = p[p$key == 'Estimated',],
              aes(x = date, ymin = LCI, ymax =  UCI),
              fill = 'gray', alpha = 0.5) +
  geom_line(data = p[p$key %in% c('Observed', 'Estimated'),],
            aes(x = date, y = val, linetype = key)) +
  labs(x=NULL, y = 'Estimated D50', linetype = NULL) +
  facet_wrap(~array, nrow = 2) +
  theme_bw()
```


