---
title: "Gammin'"
output: html_notebook
---
```{r setup, echo = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Let's GAM it up
```{r}
library(mgcv)
data <- readRDS('data and imports/rangetest_no_outliers.RDS')
```

Compare model with and without random array term.
```{r}
f_mod <- gamm(d50_adj ~ s(average_temperature, bs = 'cr') +
            s(average_noise, bs = 'cr') +
            s(tilt_angle, bs = 'cr'),
            method = 'REML',
         data = data)


r_mod <- gamm(d50_adj ~ s(average_temperature, bs = 'cr') +
            s(average_noise, bs = 'cr') +
            s(tilt_angle, bs = 'cr'),
         random = list(array = ~ 1),
            method = 'REML',
         data = data)


anova(f_mod$lme, r_mod$lme)

```

**need to update this. random might be needed after tilt adjustment**

This is barely significant when using REML, and not significant if using other methods. There's probably no advantage to including random array term. Refit using `gam`.
```{r}
mod <- gam(d50_adj ~ s(average_temperature, bs = 'cr') +
            s(average_noise, bs = 'cr') +
            s(tilt_angle, bs = 'cr'),
           method = 'REML',
         data = data, select = T)
summary(mod)
plot(mod, pages = 1, shade = T)
```
```{r}
gam.check(mod)
```


Keep chugging. Try different error structure? I'm choosing the [Gamma distribution](https://en.wikipedia.org/wiki/Gamma_distribution) because:
*  Response variable (D50) cannot be negative
*  The variance of the response variable seems to increase with increasing mean.
```{r}
mod_gamma <- gam(d50_adj ~ s(average_temperature, bs = 'cs') +
            s(average_noise, bs = 'cs') +
            s(tilt_angle, bs = 'cs'),
            family = Gamma(),
            method = 'REML',
         data = data, select = T)
summary(mod_gamma)
plot(mod_gamma)
```

```{r}
AIC(mod_gamma)
gam.check(mod_gamma)
```


```{r}
vis.gam(mod_gamma, type = 'response', theta = 130)
```

Okay, I really want to see if I can plot the smooths on the response scale. I'm following the instructions on [this Cross Validated answer](https://stats.stackexchange.com/questions/166553/creating-marginal-plots-when-using-mgcv-gam-package-in-r). First up, I'm checking temperature. Keeping noise at its median and tilt at its mean.

```{r}



bwt_fit <- data.frame(average_temperature = seq(min(data$average_temperature),
                                                max(data$average_temperature),
                                                length = 100),
                      average_noise = median(data$average_noise),
                      tilt_angle = mean(data$tilt_angle))
bwt_smooth <- predict(mod_gamma, bwt_fit, type = 'response', se.fit = T)


plot(bwt_fit$average_temperature, bwt_smooth$fit, type = 'l')
lines(bwt_fit$average_temperature, bwt_smooth$fit + 2 * bwt_smooth$se.fit, lty = 2)
lines(bwt_fit$average_temperature, bwt_smooth$fit - 2 * bwt_smooth$se.fit, lty = 2)
```

That kinda worked. Try noise with BWT and tilt at their mean.
```{r}
noise_fit <- data.frame(average_temperature = mean(data$average_temperature),
                      average_noise = seq(min(data$average_noise),
                                          max(data$average_noise),
                                          length = 100),
                      tilt_angle = mean(data$tilt_angle))
noise_smooth <- predict(mod_gamma, noise_fit, type = 'response', se.fit = T)


plot(noise_fit$average_noise, noise_smooth$fit, type = 'l')
lines(noise_fit$average_noise, noise_smooth$fit + 2 * noise_smooth$se.fit, lty = 2)
lines(noise_fit$average_noise, noise_smooth$fit - 2 * noise_smooth$se.fit, lty = 2)
```

Now tilt; BWT at mean, noise at median. 
```{r}
tilt_fit <- data.frame(average_temperature = mean(data$average_temperature),
                       average_noise = median(data$average_noise),
                       tilt_angle = seq(min(data$tilt_angle),
                                        max(data$tilt_angle),
                                        length = 100))
tilt_smooth <- predict(mod_gamma, tilt_fit, type = 'response', se.fit = T)


plot(tilt_fit$tilt_angle, tilt_smooth$fit, type = 'l')
lines(tilt_fit$tilt_angle, tilt_smooth$fit + 2 * tilt_smooth$se.fit, lty = 2)
lines(tilt_fit$tilt_angle, tilt_smooth$fit - 2 * tilt_smooth$se.fit, lty = 2)
```


```{r}
mod_gamma_mix <- gam(d50_adj ~ s(average_temperature, bs = 'cs',
                                      by  = as.factor(array)) +
            s(average_noise, bs = 'cs',
                                      by  = as.factor(array)) +
            s(tilt_angle, bs = 'cs',
                                      by  = as.factor(array)),
            family = Gamma(),
            method = 'REML',
         data = data, select = T)
summary(mod_gamma_mix)
```
