---
title: "Removing outliers from D50 time series"
output:
  html_document:
    df_print: paged
---

```{r setup, echo = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r, warning = F, message = F}
library(tsoutliers)
library(dplyr)

# Read in data
data <- readRDS('data and imports/rangetest_logit_data.rds')
names(data) <- tolower(gsub(' ', '_', names(data)))

# D50 time series calculated from logit modeling:
d50_wea <- ts(na.omit(data[data$array == 'MD WEA', 'd50']))
d50_inner <- ts(na.omit(data[data$array == 'Inner', 'd50']))
```

There seem to be some outliers, mostly in the WEA series.

```{r}
plot(cbind(d50_inner, d50_wea), main = NA)
```

I want to remove these before using predictive models (GAMs). Using the default cval (3.45) identifies far more outliers than I feel are realistic.

```{r}
tso_wea <- tso(d50_wea, types = 'AO')
plot(tso_wea)
```

I believe that part of this is due to the clear level shift at 152.

```{r, warning = F}
tso(d50_wea, types = 'LS')
```

If we just take the series after the shift, only four outliers are detected.

```{r}
wea_shift <- ts(d50_wea[152:230])
wea_shift <- tso(wea_shift, types = 'AO')
plot(wea_shift)
```


Because of this, I'm bumping the cval to 4.
```{r}
tso_wea <- tso(d50_wea, types = 'AO', cval = 4)
plot(tso_wea)

tso_inner <- tso(d50_inner, types = 'AO', cval = 4)
plot(tso_inner)
```

Better. Adjust the original data.

```{r}
data <- data %>%
  filter(!is.na(d50)) %>% 
  arrange(array) %>% 
  mutate(d50_adjust = c(tso_inner$yadj, tso_wea$yadj))

saveRDS(data, 'data and imports/rangetest_logit_data_adj.RDS')
```

